Replication in Archaeological Science: Shape as a Measure of Weapon Standardisation

### 2020-02-25 analysis/paper.Rmd
code cleaning and reorg

<insert load packages block here>

@@

<insert set working dir here>

@@

<insert read in data here>

@@

```{r load packages, include=TRUE}
#adding this line as a git test
#install specific versions of pakages called out in the paper
#require(devtools)
library(devtools)
#install_version("geomorph", version = "3.0.6", repos = "http://cran.rstudio.com")
#install_version("geomorph", version = "3.0.6", repos = "http://cran.r-project.org")
#install.packages("http://https://cran.r-project.org/src/contrib/Archive/geomorph/geomorph_3.0.6.tar.gz", repos=NULL, type="source")
library(geomorph) # geomorph 3.0.6

install_version("shapes", version = "1.2.3", repos = "http://cran.rstudio.com/") #"http://cran.rstudio.com")
#install_version("shapes", version = "1.2.3", repos = "http://cran.rstudio.com/") #"http://cran.us.r-project.org")
library(shapes) # shapes 1.2.3

#package to support updating geomoreph code
library(RRPP)

```{r load data, include=TRUE}
# Read the datafile "lances_M" containing a matrix of dimensional data (metric)

#replication file names did not survive publication!
#lances_M<-read.csv("Data/1-s2.0-S0305440318306757-mmc2.csv",header=TRUE)
lances_M<-read.csv(here("Data","1-s2.0-S0305440318306757-mmc2.csv"),header=TRUE)

```{r analysis 1, include=TRUE}
# Before undertaking any descriptive statistics of the metric dimensions of the lances, it is important to re-sort the data matrix and subset the data.
lances_M$site<-factor(lances_M$site,
                    levels = c("Ejsbol","Nydam", "Illerup"),
                    labels = c("Ejsbol","Nydam", "Illerup"),
                    ordered=T)                                # This simply converts the variable 'site' into an ordered variable of factors with levels.

# Principal Components Analysis of Havor lance dimensions
totalpca<-prcomp(~total_length+blade_length+socket_length+sock_bdiam+sock_sdiam+width+thickness,data=total,center=T,scale=T)
totalpca        # By typing 'totalpca' a full summary of the components is printed in the terminal.
str(totalpca)   # More information can be viewed about the PCA results by looking at the 'total' object in more detail.

# Heirarchical Cluster Analysis of Havor lance dimensions
#totalhc <-hclust(dist(total[,(c(1,5:11))]), "average")  # The cluster analysis is performed
#totalhc1<-as.dendrogram(totalhc)                        # The cluster analysis results are converted into a dendrogram object
#plot(totalhc1)                                          # Plots the resulting dendrogram object
```
```{r analysis 1.1, include=FALSE}
# Heirarchical Cluster Analysis of Havor lance dimensions
# The cluster analysis is performed
totalhc <-hclust(dist(total[,(c(1,5:11))]), "average")
# The cluster analysis results are converted into a dendrogram object
totalhc1<-as.dendrogram(totalhc)                      
```
```{r analysis 1.2, include=TRUE}
# Plots the resulting dendrogram object
plot(totalhc1)
```
```{r analysis 2, include=FALSE}

                          #-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-
                          #### 4. GEOMETRIC MORPHOMETRICS (GMM): LANDMARKS DATA ----
                          #-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-

# Create the links (this specifies the order of landmarks by which to connect linking lines)
outline<-rbind(c(1,2),c(2,3),c(3,4),c(4,5),c(5,6),c(6,7),c(7,1))

                          #-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-
                          #### 5. GMM DATA MANIPULATION ----
                          #-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-

# Subset the data into different sites
Eco<-lances_geo[,,1:6]    # Ejsbol (n=6)
Nco<-lances_geo[,,7:30]   # Nydam (n=24)
Ico<-lances_geo[,,31:78]  # Illerup (n=48)

```{r plotting digitized specs, include=FALSE}
                          #-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#
                          #### 6. PLOTTING DIGITISED SPECIMENS ----
                          #-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#

# This function plots the landmarks of specimens recorded (as digitised)
plotAllSpecimens(lances_geo,links=outline)  # Plot landmarks of all specimens (digitised)
plotAllSpecimens(Eco,links=outline)         # Plot landmarks of Ejsbol lances (digitised)
plotAllSpecimens(Nco,links=outline)         # Plot landmarks of Nydam lances (digitised)
plotAllSpecimens(Ico,links=outline)         # Plot landmarks of Illerup lances (digitised)

```{r GPA, include=FALSE}

                          #-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-
                          #### 7. GENERALISED PROCRUSTES ANALYSIS (GPA) ----
                          #-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-

# GPA is performed on the lances, including the semi-landmarks at the socket.
# The 'curves' argument is ignored to have permanent landmarks
lancespr<-gpagen(lances_geo)  # GPA of all lances
Epr<-gpagen(Eco)              # GPA of Ejsbol lances
Npr<-gpagen(Nco)              # GPA of Nydam lances
Ipr<-gpagen(Ico)              # GPA of Illerup lances

```{r plotting GPA, include=FALSE}

                          #-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#
                          #### 8. PLOTTING GPA SPECIMENS ----
                          #-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#

# This function plots the landmarks of GPA specimens
plotAllSpecimens(lancespr$coords,links=outline) # Plot landmarks of all specimens (GPA)
plotAllSpecimens(Epr$coords,links=outline)      # Plot landmarks of Ejsbol lances (GPA)
plotAllSpecimens(Npr$coords,links=outline)      # Plot landmarks of Nydam lances (GPA)
plotAllSpecimens(Ipr$coords,links=outline)      # Plot landmarks of Illerup lances (GPA)

# This function plots the shapes of GPA specimens
plotshapes(lancespr$coords,joinline=c(1:7,1),symbol=NA) # Plot shapes of all specimens (GPA)
plotshapes(Epr$coords,joinline=c(1:7,1),symbol=NA)      # Plot shapes of Ejsbol lances (GPA)
plotshapes(Npr$coords,joinline=c(1:7,1),symbol=NA)      # Plot shapes of Nydam lances (GPA)
plotshapes(Ipr$coords,joinline=c(1:7,1),symbol=NA)      # Plot shapes of Illerup lances (GPA)

```{r check size vs shape, include=FALSE}
                          #-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-
                          #### 9. ALLOMETRY: CHECKING IF SHAPE IS RELATED TO SIZE ----
                          #-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-

# Create a geomorph data frame for analysis
lances_geomorph<-geomorph.data.frame(shape=lancespr$coords,site=site,size=lancespr$Csize,ind=lances_GMM$site_lance.id) # Parse the data into a geomorph data frame

# We can test to see if there is an association between centroid size and shape, before conducting further statistical analyses.

# The results are significant, but they cannot be explained by site difference
lances_ANOVA<-procD.allometry(shape ~ size, ~site, data=lances_geomorph,method="PredLine") # Optional right-hand formula for the inclusion of groups; "~site" can be removed to perform ANOVA on whole dataset
?procD.allometry
# lances_ANOVA<-procD.allometry(shape ~ size, data=lances_geomorph,method="PredLine") # As above, but with "~site" removed
#plot(lances_ANOVA)
#print(lances_ANOVA)

#Call:
#  procD.allometry(f1 = shape ~ size, f2 = ~site, data = lances_geomorph,      method = "PredLine") 
#
#Homogeneity of Slopes Test
#Df     RSS       SS      Rsq      F       Z Pr(>F)
#Common Allometry  74 0.76683                                        
#Group Allometries 72 0.74137 0.025466 0.030754 1.2366 0.67153  0.246
#
#The null hypothesis of parallel slopes is supported
#based on a significance criterion of alpha = 0.05 
#
#Based on the results of this test, the following ANOVA table is most appropriate
#
#Type I (Sequential) Sums of Squares and Cross-products
#Randomized Residual Permutation Procedure Used
#1000 Permutations
#ANOVA effect sizes and P-values based on empirical F distributions
#
#Df      SS       MS     Rsq      F        Z Pr(>F)  
#log(size)  1 0.05502 0.055024 0.06645 5.3098  1.91423  0.015 *
#  site       2 0.00619 0.003096 0.00748 0.2987 -0.80096  0.796  
#Residuals 74 0.76683 0.010363 0.92607                         
#Total     77 0.82805                                          
#---
#  Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

```{r test mean shape diff, include=FALSE}
                          #-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-
                          #### 10. TESTING FOR MEAN SHAPE DIFFERENCE: HOTELLING T^2 STATISTIC AND GOODALL F STATISTIC ----
                          #-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-

# For testing the difference in mean shape (between sites) we use the raw digitised coordinates.
# By using 'scale=TRUE' in our argument, procrustes (with scaling) is carried out in the procedure.

test1<-testmeanshapes(Eco,Nco,resamples=1000,replace=TRUE,scale=TRUE) # Comparing the mean shape of Ejsbol with Nydam
# Results:
# Hotelling T^2 statistic: F=3.595 P=0.099
# Goodall F statistic: F= 0.084 P= 0.806

test2<-testmeanshapes(Eco,Ico,resamples=1000,replace=TRUE,scale=TRUE) # Comparing the mean shape of Ejsbol with Illerup
# Results:
# Hotelling T^2 statistic: F=5.746 P=0.004
# Goodall F statistic: F= 0.752 P= 0.430

test3<-testmeanshapes(Ico,Nco,resamples=1000,replace=TRUE,scale=TRUE) # Comparing the mean shape of Illerup with Nydam
# Results:
# Hotelling T^2 statistic: F=0.713 P=0.739
# Goodall F statistic: F= 0.829 P= 0.490

```{r ANOVA test, include=FALSE}
                          #-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#
                          #### 11. PROCRUSTES ANALYSIS OF VARIANCE (ANOVA) ----
                          #-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#

# The Procrustes ANOVA is identical to permutational-MANOVA.
# We want to perform statistical tests for mean shape difference on the GPA specimens.

# Procustres ANOVA
lances_ANOVA2<-procD.lm(shape~site,data=lances_geomorph, iter=99)
print(lances_ANOVA2)

#Call:
#  procD.lm(f1 = shape ~ site, iter = 99, data = lances_geomorph)
#
#Type I (Sequential) Sums of Squares and Cross-products
#Randomized Residual Permutation Procedure Used
#100 Permutations
#ANOVA effect sizes and P-values based on empirical F distributions
#
#Df      SS        MS     Rsq      F        Z Pr(>F)
#site       2 0.01286 0.0064307 0.01553 0.5917 -0.44955    0.7
#Residuals 75 0.81519 0.0108691 0.98447                       
#Total     77 0.82805       
```

```{r PCA, include=FALSE}

                          #-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#
                          #### 12. PRINCIPAL COMPONENTS ANALYSIS (PCA) OF GENERALISED PROCRUSTES ----
                          #-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#

# PCA is carried out on procrustes aligned specimens.
lancespca<-plotTangentSpace(lancespr$coords,label=TRUE)
lancespca       # By typing 'lancespca' a full summary of the components is printed in the terminal.
str(lancespca)  # More information can be viewed about the PCA results by looking at the 'lancespca' object in more detail.
```

```{r JCA, include=FALSE}
                          #-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#
                          #### 13. HEIRARCHICAL CLUSTER ANALYSIS (HCA) OF GENERALISED PROCRUSTES ----
                          #-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#

# HCA is carried out on procustes aligned specimens.
hc<-hclust(dist(two.d.array(lancespr$coords)), "average")  # The cluster analysis is performed
hc1<-as.dendrogram(hc)                  # The cluster analysis results are converted into a dendrogram object
plot(hc1)                               # Plots the resulting dendrogram object

```{r bilateral symmetry, include=FALSE}

                          #-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-
                          #### 14. BILATERAL SYMMETRY INVESTIGATION ----
                          #-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-#-

# Define a matrix of the paired landmarks (i.e. S1 and S2, N1 and N2, W1 and W2)
lances_land.pairs<-matrix(c(1,7,2,6,3,5),nrow=3,ncol=2,byrow=TRUE)

# Perform an analysis f the directional and fluctuating asymmetry of Havor lances
lance.sym<-bilat.symmetry(A=shape,ind=ind,object.sym=TRUE,land.pairs=lances_land.pairs,data=lances_geomorph,RRPP=FALSE,iter=999)

# View the results (ANOVA table)
lance.sym

#Call:
#  bilat.symmetry(A = shape, ind = ind, object.sym = TRUE, land.pairs = lances_land.pairs,  
#                 data = lances_geomorph, iter = 999, RRPP = FALSE) 
#
#Symmetry (data) type: Object 
#
#Type I (Sequential) Sums of Squares and Cross-products
#Randomization of Raw Values used
#1000 Permutations
#
#Shape ANOVA
#Df      SS        MS     Rsq        F       Z Pr(>F)   
#ind        77 1.55153 0.0201498 0.99820 577.3822 23.2156  0.001 **
#  side        1 0.00011 0.0001059 0.00007   3.0349  1.3652  0.063 . 
#Residuals  77 0.00269 0.0000349 0.00173                           
#Total     155 1.55433                                             
#---
#  Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

# Plot the results
plot(lance.sym, warpgrids=TRUE)
```

@@

#COMMENTED OUT BY THOMAS IN ORDER TO KNIT
#ggpairs(lances_M, columns = c("total_length", "blade_length", "socket_length", #"sock_bdiam", "sock_sdiam", "width", "thickness"), 
#        upper=list(continuous = upper_fn),
#        diag=list(continuous = diag_fn),
#        lower=list(continuous = "cor")) + theme_minimal() +
#        ggsave(here("Images","replication_fig_8v1.png"))

#Copied to try to adjust to get to knit in Thomas's env
#COMMENTED OUT BY THOMAS IN ORDER TO KNIT
#ggpairs(lances_M, columns = c('total_length', 'blade_length', 'socket_length', 'sock_bdiam', 'sock_sdiam', 'width', #'thickness'),  
#        upper=list(continuous = upper_fn),
#        diag=list(continuous = diag_fn),
#        lower=list(continuous = "cor")) + theme_minimal() +
#        ggsave(here("Images","replication_fig_8v1.png"))

@@

```{r replication corr test 2, include=TRUE, echo=FALSE}
#COMMENTED OUT BY THOMAS IN ORDER TO KNIT
#ggpairs(lances_M[complete.cases(lances_M),], columns = c("total_length", "blade_length", #"socket_length", "sock_bdiam", "sock_sdiam", "width", "thickness"), 
#        upper=list(continuous = upper_fn),
#        diag=list(continuous = diag_fn),
#        lower=list(continuous = "cor")) + theme_minimal() +
#        ggsave(here("Images","replication_fig_8_complete_cases_manual.png"))

#COMMENTED OUT BY THOMAS IN ORDER TO KNIT
#ggpairs(blade, columns = c("total_length", "blade_length", "socket_length", #"sock_bdiam", "sock_sdiam", "width", "thickness"), 
#        upper=list(continuous = upper_fn),
#        diag=list(continuous = diag_fn),
#        lower=list(continuous = "cor")) + theme_minimal() +
#        ggsave(here("Images","replication_fig_8_blade.png"))

#COMMENTED OUT BY THOMAS IN ORDER TO KNIT
#ggpairs(socket, columns = c("total_length", "blade_length", "socket_length", #"sock_bdiam", "sock_sdiam", "width", "thickness"), 
#        upper=list(continuous = upper_fn),
#        diag=list(continuous = diag_fn),
#        lower=list(continuous = "cor")) + theme_minimal() +
#        ggsave(here("Images","replication_fig_8_socket.png"))

#COMMENTED OUT BY THOMAS IN ORDER TO KNIT
#ggpairs(total, columns = c("total_length", "blade_length", "socket_length", #"sock_bdiam", "sock_sdiam", "width", "thickness"), 
#        upper=list(continuous = upper_fn),
#        diag=list(continuous = diag_fn),
#        lower=list(continuous = "cor")) + theme_minimal() +
#        ggsave(here("Images","replication_fig_8_blades_and_sockets.png"))
```

>>>>

```{r load packages, include=TRUE}
#adding this line as a git test
#install specific versions of packages called out in the paper
#require(devtools)
library(devtools)
#install_version("geomorph", version = "3.0.6", repos = "http://cran.rstudio.com")
#install_version("geomorph", version = "3.0.6", repos = "http://cran.r-project.org")
#install.packages("http://https://cran.r-project.org/src/contrib/Archive/geomorph/geomorph_3.0.6.tar.gz", repos=NULL, type="source")
library(geomorph) # geomorph 3.0.6

#install_version("shapes", version = "1.2.3", repos = "http://cran.rstudio.com/") #"http://cran.rstudio.com")
#install_version("shapes", version = "1.2.3", repos = "http://cran.rstudio.com/") #"http://cran.us.r-project.org")
library(shapes) # shapes 1.2.3

#package to support updating geomorph code for those that cannot use old package
#library(RRPP)

@@

@@

```{r load data, include=TRUE}
# Read the appendix datafile containing a matrix of dimensional data (metric) as "lances_M" 
lances_M<-read.csv(here("Data","1-s2.0-S0305440318306757-mmc2.csv"),header=TRUE)

```{r analysis 1, include=TRUE}

# CODE PROVIDED BY AUTHORS IN TECHNICAL APPENDIX

# Before undertaking any descriptive statistics of the metric dimensions of the lances, it is important to re-sort the data matrix and subset the data.
lances_M$site<-factor(lances_M$site,
                    levels = c("Ejsbol","Nydam", "Illerup"),
                    labels = c("Ejsbol","Nydam", "Illerup"),
                    ordered=T)                                # This simply converts the variable 'site' into an ordered variable of factors with levels.

# GPA is performed on the lances, including the semi-landmarks at the socket.
lancespr<-gpagen(lances_geo)  # GPA of all lances

# Create a geomorph data frame for analysis
lances_geomorph<-geomorph.data.frame(shape=lancespr$coords,site=site,size=lancespr$Csize,ind=lances_GMM$site_lance.id) # Parse the data into a geomorph data frame
```

@@

@@

#COMMENTED OUT BY THOMAS IN ORDER TO KNIT
ggpairs(lances_M, columns = c("total_length", "blade_length", "socket_length", "sock_bdiam", "sock_sdiam", "width", "thickness"), 
        upper=list(continuous = upper_fn),
        diag=list(continuous = diag_fn),
        lower=list(continuous = "cor")) + theme_minimal() #+
        #ggsave(here("Images","replication_fig_8v1.png"))

#Copied to try to adjust to get to knit in Thomas's env
#COMMENTED OUT BY THOMAS IN ORDER TO KNIT
ggpairs(lances_M, columns = c('total_length', 'blade_length', 'socket_length', 'sock_bdiam', 'sock_sdiam', 'width', 'thickness'),  
        upper=list(continuous = upper_fn),
        diag=list(continuous = diag_fn),
        lower=list(continuous = "cor")) + theme_minimal() #+
        #ggsave(here("Images","replication_fig_8v1.png"))

@@

```{r replication corr test 2, include=TRUE, echo=FALSE}
#COMMENTED OUT BY THOMAS IN ORDER TO KNIT
ggpairs(lances_M[complete.cases(lances_M),], columns = c("total_length", "blade_length", "socket_length", "sock_bdiam", "sock_sdiam", "width", "thickness"), 
        upper=list(continuous = upper_fn),
        diag=list(continuous = diag_fn),
        lower=list(continuous = "cor")) + theme_minimal() #+
#        ggsave(here("Images","replication_fig_8_complete_cases_manual.png"))

#COMMENTED OUT BY THOMAS IN ORDER TO KNIT
ggpairs(blade, columns = c("total_length", "blade_length", "socket_length", "sock_bdiam", "sock_sdiam", "width", "thickness"), 
        upper=list(continuous = upper_fn),
        diag=list(continuous = diag_fn),
        lower=list(continuous = "cor")) + theme_minimal() #+
#        ggsave(here("Images","replication_fig_8_blade.png"))

#COMMENTED OUT BY THOMAS IN ORDER TO KNIT
ggpairs(socket, columns = c("total_length", "blade_length", "socket_length", "sock_bdiam", "sock_sdiam", "width", "thickness"), 
        upper=list(continuous = upper_fn),
        diag=list(continuous = diag_fn),
        lower=list(continuous = "cor")) + theme_minimal() #+
#        ggsave(here("Images","replication_fig_8_socket.png"))

#COMMENTED OUT BY THOMAS IN ORDER TO KNIT
ggpairs(total, columns = c("total_length", "blade_length", "socket_length", "sock_bdiam", "sock_sdiam", "width", "thickness"), 
        upper=list(continuous = upper_fn),
        diag=list(continuous = diag_fn),
        lower=list(continuous = "cor")) + theme_minimal() #+
#        ggsave(here("Images","replication_fig_8_blades_and_sockets.png"))
```
